---
title: "Covid 19 analysis"
author: "Jie Shen"
date: "2023-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## File and Data

This is a R Markdown document for **COVID 19 project for China**. The data used in this project can be found at "https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series". Please visit the site for detailed data description.


## Project goal
The project is to discover patterns and trends from Covid data in China. I want to explore things like the Covid cases and deaths trends over the years, and what states are best and worst.

## Packages needed
Be sure the following packages are installed first:

* tidyverse
* ggplot2

## Load Packages
```{r load packages, results="hide",message=FALSE}

library(tidyverse)
library(ggplot2)
library(forcats)
library(lubridate)
```

## Import Data and clean up
```{r import data}
#Import data from webnsite
url_in<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names<-c("time_series_covid19_confirmed_global.csv","time_series_covid19_deaths_global.csv")

urls=str_c(url_in, file_names)
global_cases<-read_csv(urls[1])
global_deaths<-read_csv(urls[2])
```
Now let's take a look and do some clean up
```{r global cases and deaths}
# Take a look
head(global_cases)
head(global_deaths)

# Need to pivot dates to rows
global_cases<-global_cases %>%
  pivot_longer(cols= -c("Province/State", "Country/Region", Lat, Long),
                       names_to="date",
                       values_to="cases")
head(global_cases)

# Do similar things to global deaths
global_deaths<-global_deaths %>%
  pivot_longer(cols= -c("Province/State", "Country/Region", Lat, Long),
                       names_to="date",
                       values_to="deaths")

# Combine global cases and deaths
global<- global_cases %>%
        full_join(global_deaths) %>%
        mutate(date=mdy(date)) %>%
        rename(Country_Region='Country/Region',
               Province_State ='Province/State')

# Take a look again
head(global)

# US data has "Combined_Key". Add this to global data too.
global<-global%>%
  unite("Combined_Key",
       c("Province_State", "Country_Region"),
       sep=", ",
       na.rm=TRUE,
       remove=FALSE
  )


# US data has "Combined_Key". Add this to global data too.
global<-global%>%
  unite("Combined_Key",
       c("Province_State", "Country_Region"),
       sep=", ",
       na.rm=TRUE,
       remove=FALSE
  )

# Take another look
head(global)

# Summary statistics
summary(global)

```

We can see the earliest date is 2020-01-22 and the latest is 2023-03-09.

Since tt's unfair to compare the numbers from big population state to a small state, I also want to see cases and deaths per populations. I found the population data set on the same github website.

```{r get population}
# Import population data
uid_lookup_url="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid=read_csv(uid_lookup_url)
# After looking through the columns, exclude unwanted columns %>%
uid<-uid%>% select(-c(Lat, Long_, Combined_Key, iso2,  iso3,  code3,Admin2, UID, FIPS) )

# Add population column to global data
global<-global%>%
  full_join(uid, by=c("Province_State", "Country_Region"))

# Take another look
head(global)
```
## Analysis
### Get per state and total Country numbers
```{r China by state}
# Get a China data frame
CN<-global%>%filter(Country_Region=="China")

# China by state total cases, deaths, and death per million population
CN_by_state<-CN%>%
  group_by( Country_Region,Province_State, date) %>%
  summarise(cases=sum(cases), deaths=sum(deaths), Population = sum(Population)) %>%
  mutate(death_per_mill = deaths/Population*1000000) %>%
  ungroup()

#Take a look
tail(CN_by_state)

# China Totals
CN_totals<- CN%>%
  group_by( Country_Region, date) %>%
  summarise(cases=sum(cases), deaths=sum(deaths), Population = sum(Population)) %>%
  mutate(death_per_mill = deaths/Population*1000000) %>%
  arrange(death_per_mill) %>%
 ungroup()

#Take a look
tail(CN_totals)
```

### Visualization CN totals

```{r CN Totals Viz}
# Visualize CN totals
options(repr.plot.width=30, repr.plot.height=10)
CN_totals %>%
  filter(cases>0) %>%
  ggplot(aes(x=date, y=cases))  +
  geom_line(aes(color="cases")) +
  geom_point(aes(color="cases")) +
  geom_line(aes(y=deaths, color="deaths")) +
  geom_point(aes(y=deaths, color="deaths")) +
  scale_y_log10() +
   scale_x_date(date_labels = "%y-%b", date_breaks = "4 month") +
  theme(legend.position='bottom', axis.text=element_text(angle=90, size=10)) +
  labs(title="COVID 19 in China - total cases and deaths", y=NULL)
```

## How about new cases and new deaths?
When looking at trends, it's good to see how many new cases and new deaths. Let's add those columns
```{r new cases and deaths}
# Add new cases columns to China data
CN_by_state<- CN_by_state%>% arrange(Country_Region, Province_State, date) %>%
      mutate(new_cases=cases-lag(cases), new_deaths=deaths-lag(deaths))
    
CN_totals<- CN_totals%>% arrange(Country_Region, date) %>%
      mutate(new_cases=cases-lag(cases), new_deaths=deaths-lag(deaths))
    
# Take a look
tail(CN_by_state)
tail(CN_totals)
```
## Visualize new cases and deaths in China
```{r vis new cases and deaths}
# Visualize China totals
options(repr.plot.width=30, repr.plot.height=10)
CN_totals %>%
  filter(cases>0) %>%
  ggplot(aes(x=date, y=new_cases))  +
  geom_line(aes(color="new_cases")) +
  geom_point(aes(color="new_cases")) +
  geom_line(aes(y=deaths, color="new_deaths")) +
  geom_point(aes(y=deaths, color="new_deaths")) +
  scale_y_log10() +
  scale_x_date(date_labels = "%y-%b", date_breaks = "4 month") +
  theme(legend.position='bottom', axis.text=element_text(angle=90, size=10)) +
  labs(title="COVID 19 in China - new cases and deaths", y=NULL)
```
## What are the worst and best states in China?

### CN by states
Let's see which states are best/worst (in term of death/population)
```{r worst best states}
CN_state_totals <- CN_by_state %>%
   group_by(Province_State) %>%
   summarize(cases=max(cases),
       deaths= max(deaths),
       Population=max(Population),
       cases_per_thou=1000*cases/Population,
       deaths_per_thou=1000*deaths/Population)
CN_state_totals %>% slice_min(deaths_per_thou,n=10)
CN_state_totals %>% slice_max(deaths_per_thou,n=10)
```

## visualize state of interest
I want to visualize the top 3 worst states
```{r top 3 viz}
state<- "Hong Kong"
CN_by_state %>%
  filter(Province_State==state) %>%
  ggplot(aes(x=date, y=cases))  +
  geom_line(aes(color="cases")) +
  geom_point(aes(color="cases")) +
  geom_line(aes(y=deaths, color="deaths")) +
  geom_point(aes(y=deaths, color="deaths")) +
  scale_y_log10() +
  scale_x_date(date_labels = "%y-%b", date_breaks = "4 month") +
  theme(legend.position='bottom', axis.text=element_text(angle=90, size=10)) +
  labs(title=str_c("COVID 19 in ", state," - total cases and deaths"), y=NULL)

state<- "Macau"
CN_by_state %>%
  filter(Province_State==state) %>%
  ggplot(aes(x=date, y=cases))  +
  geom_line(aes(color="cases")) +
  geom_point(aes(color="cases")) +
  geom_line(aes(y=deaths, color="deaths")) +
  geom_point(aes(y=deaths, color="deaths")) +
  scale_y_log10() +
  scale_x_date(date_labels = "%y-%b", date_breaks = "4 month") +
  theme(legend.position='bottom', axis.text=element_text(angle=90, size=10)) +
  labs(title=str_c("COVID 19 in ", state," - total cases and deaths"), y=NULL)

state<- "Hubei"
CN_by_state %>%
  filter(Province_State==state) %>%
  ggplot(aes(x=date, y=cases))  +
  geom_line(aes(color="cases")) +
  geom_point(aes(color="cases")) +
  geom_line(aes(y=deaths, color="deaths")) +
  geom_point(aes(y=deaths, color="deaths")) +
  scale_y_log10() +
  scale_x_date(date_labels = "%y-%b", date_breaks = "4 month") +
  theme(legend.position='bottom', axis.text=element_text(angle=90, size=10)) +
  labs(title=str_c("COVID 19 in ", state," - total cases and deaths"), y=NULL)

